---
title: "Using ylistjp"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using ylistjp}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

`ylistjp` is an unofficial R package for looking up scientific plant names from
Japanese plant names using the Vascular Plant Japanese Name Checklist ver. 1.10.

The checklist is useful when your starting point is a Japanese plant name, such
as `コナラ`, and you want the corresponding scientific name in a reproducible R
workflow.

## Installation

Install the package from GitHub:

```{r}
# install.packages("pak")
pak::pak("maple60/ylistjp")
```

Then load it:

```{r}
library(ylistjp)
```

## First Lookup

Use `scientific_name()` when you have a Japanese plant name and want the standard
scientific name.

```{r}
scientific_name("コナラ")
#> [1] "Quercus serrata"
```

If you want the author string as recorded in the checklist, set
`with_author = TRUE`.

```{r}
scientific_name("コナラ", with_author = TRUE)
#> [1] "Quercus serrata Murray"
```

The function is vectorized:

```{r}
scientific_name(c("コナラ", "ミズナラ"))
```

## How the Data Is Loaded

`ylistjp` does not bundle or redistribute checklist data. The first call that
needs the data downloads the checklist Excel file and stores it in the user's
local R cache.

You can also download or refresh the file explicitly:

```{r}
japanese_name_download()
japanese_name_download(overwrite = TRUE)
```

Load the cached data as a regular data frame:

```{r}
checklist <- japanese_name_load()
head(checklist)
```

To force a refresh before loading:

```{r}
checklist <- japanese_name_load(refresh = TRUE)
```

## Exact Lookup Behavior

`scientific_name()` is intentionally conservative:

- it exact-matches the checklist `和名` and `別名` columns;
- it only uses rows where `ステータス` is `標準`;
- it returns `NA_character_` when there is no standard exact match;
- it errors if more than one standard exact match is found, so you can inspect
  the candidates manually.

This behavior is designed for scripts and analyses where silent fuzzy matching
would be risky.

## Searching Candidates

Use `japanese_name_search()` when you want to inspect possible matches.

```{r}
japanese_name_search("コナラ")
```

By default, `japanese_name_search()` searches Japanese names with partial matching.
You can search specific fields:

```{r}
japanese_name_search("Quercus serrata", field = "scientific")
japanese_name_search("ナラ", field = "alias")
japanese_name_search("コナラ", field = "all")
```

Use exact matching when you want only exact candidate rows:

```{r}
japanese_name_search("コナラ", exact = TRUE)
```

## A Practical Workflow

A typical workflow is:

1. Use `scientific_name()` for names that should be standard and unambiguous.
2. Use `japanese_name_search()` for names that return `NA` or need manual checking.
3. Keep both the original Japanese name and returned scientific name in your
   analysis table.
4. Cite the checklist when you use checklist-based names in a report, paper, or dataset.

Example:

```{r}
plants <- data.frame(
  japanese_name = c("コナラ", "ミズナラ", "存在しない植物名"),
  stringsAsFactors = FALSE
)

plants$scientific_name <- scientific_name(plants$japanese_name)
plants
```

## Checking Names with GBIF

`gbif_match()` is a small optional helper around the GBIF species match API.
Use it after obtaining a scientific name from the checklist when you want to
check how the name is represented in an international biodiversity data source.

```{r}
gbif_match("Quercus serrata")
```

This function requires the suggested package `jsonlite`.

## Data Source and Citation

This package uses the Vascular Plant Japanese Name Checklist ver. 1.10 as the
lookup source. The checklist includes YList-derived/update data, but this
package is not affiliated with or endorsed by JBIF, YList, or the checklist
authors.

When using checklist-based results, cite the checklist:

> Yamanouchi, T., Shutoh, K., Osawa, T., Yonekura, K., Kato, S., Shiga, T.
> 2019. A checklist of Japanese plant names
> (https://gbif.jp/activities/checklist/wamei_checklist_110)

The package code is MIT licensed. Checklist data is not included in the package
and is not covered by the package license.
